---
title: "Hands-on_Ex10 - Information Dashboard Design: R methods"
author: "Hoa Nguyen Phuong "
format: html
date-modified: "last-modified" 
execute:
  echo: true 
  eval: true 
  warning: false 
  freeze: true
---

# [1]{style="color:mediumvioletred"} Objectives

-   Create bullet chart by using ggplot2

-   Create sparklines by using ggplot2

-   Build industry standard dashboard with R Shiny

# [2]{style="color:mediumvioletred"} Getting started

R packages that are used for this chapter:

-   [tidyverse](https://www.tidyverse.org/): provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

-   [lubridate](https://r4va.netlify.app/chap31): provides functions to work with dates and times more efficiently.

-   [ggthemes](https://r4va.netlify.app/chap31): is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html): provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

-   [reactable](https://glin.github.io/reactable/index.html): provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html): provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

# [3]{style="color:mediumvioletred"} The dataset

A personal database in Microsoft Access mdb format called *Coffee Chain* will be used.

```{r}
pacman::p_load(lubridate, ggthemes, reactable, reactablefmtr, gt, gtExtras, tidyverse)
```

## [3.1]{style="color:mediumvioletred"} Importing network data from files

In the code chunk below, `odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package is used used to import a database query table into R.

``` r
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

::::: goals
::: goals-header
Note
:::

::: goals-container
Before running the code chunk, R system needs to be changed to 32bit version. This is because the `odbcConnectAccess()` is based on 32bit and not 64bit. However, macOS the latest version no longer support 32-bit. We wil skip running this code, and directly use the dataset provided by the professor.
:::
:::::

## [3.4]{style="color:mediumvioletred"} Data preparation

The code chunk below is used to import *CoffeeChain.rds* into R.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

The code chunk below is used to aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

# [4]{style="color:mediumvioletred"} Bullet chart in ggplot2

The code chunk below is used to plot the bullet charts using ggplot2 functions.

```{r}
#| code-fold: true
#| code-summary: "Show code"
product_ordered <- product
product_ordered$Product <- factor(product_ordered$Product,
                                  levels = product_ordered$Product[order(product_ordered$current)])
ggplot(product_ordered, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="#F8F3D9", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="#EBE5C2", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "#504B38") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  theme(plot.background = element_rect(fill = "#f1f4f5"),
        panel.background = element_rect(fill = "#f1f4f5")) +
  coord_flip()
```

# [5]{style="color:mediumvioletred"} Plotting sparklines using ggplot2

## [5.1]{style="color:mediumvioletred"} Prepare data

First, create a sales report with monthly sales data.

```{r}
#| code-fold: true
#| code-summary: "Show code"
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

Then compute the minimum, maximum and end of the month sales.

```{r}
#| code-fold: true
#| code-summary: "Show code"
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

We calculate the 25 and 75 quantiles and add back to data frame for plotting.

```{r}
#| code-fold: true
#| code-summary: "Show code"
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

## [5.2]{style="color:mediumvioletred"} sparklines in ggplot

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = "#EDDFE0") +
  geom_line(size=0.2) +
  geom_point(data = mins, col = 'red', size = 1) +
  geom_point(data = maxs, col = 'blue', size = 1) +
  geom_text(data = mins, aes(label = Sales), vjust = -1, size = 2) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 2) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5, size = 2) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0, size = 3) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(),
        axis.text.x = element_text(size = 6, face = "bold"),
        plot.background = element_rect(fill = "#f1f4f5"),
        strip.background = element_rect(fill = "#E0F4FF") 
        ) 

```

# [6]{style="color:mediumvioletred"} Static Information Dashboard Design: gt and gtExtras methods

We will create static information dashboard by using [gt](https://gt.rstudio.com/) and [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) packages.

## [6.1]{style="color:mediumvioletred"} Plot a simple bullet chart

Use the code below to plot a bullet chart.

```{r}
#| code-fold: true
#| code-summary: "Show code"
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# [7]{style="color:mediumvioletred"} sparklines: gtExtras method

Before we can prepare the sales report by product by using gtExtras functions, code chunk below will be used to prepare the data.

```{r}
#| code-fold: true
#| code-summary: "Show code"
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

::::: goals
::: goals-header
Note
:::

::: goals-container
One of the requirement of gtExtras functions is that almost exclusively they require you to pass data.frame with list columns. In view of this, code chunk below will be used to convert the report data.frame into list columns.
:::
:::::

```{r}
#| code-fold: true
#| code-summary: "Show code"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## [7.1]{style="color:mediumvioletred"} Plot Coffechain sales report

```{r}
#| code-fold: true
#| code-summary: "Show code"
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

## [7.2]{style="color:mediumvioletred"} Add statistics

First, calculate summary statistics.

```{r}
#| code-fold: true
#| code-summary: "Show code"
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

## [7.3]{style="color:mediumvioletred"} Combine the data.frame

Next, we will add the statistics to the table.

```{r}
#| code-fold: true
#| code-summary: "Show code"
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
sales_data = left_join(sales, spark)
```

## [7.4]{style="color:mediumvioletred"} Plot the updated data.table

Now the plot has combined salse report and stats.

```{r}
#| code-fold: true
#| code-summary: "Show code"
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## [7.5]{style="color:mediumvioletred"} Combine bullet chart and sparklines

Similarly, we can combining the bullet chart and sparklines using the steps below.

```{r}
#| code-fold: true
#| code-summary: "Show code"
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()

sales_data = sales_data %>%
  left_join(bullet)

sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# [8]{style="color:mediumvioletred"} Interactive Information Dashboard Design: reactable and reactablefmtr methods

We will create interactive information dashboard by using [reactable](https://glin.github.io/reactable/index.html) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) packages. In order to build an interactive sparklines, we need to install dataui R package by using the code chunk below.

```{r}
remotes::install_github("timelyportfolio/dataui")
```

Next, we will load the package onto R environment by using the code chunk below.

```{r}
library(dataui)
```

## [8.1]{style="color:mediumvioletred"} Plotting interactive sparklines

Similar to gtExtras, to plot an interactive sparklines by using reactablefmtr package we need to prepare the list field.

```{r}
#| code-fold: true
#| code-summary: "Show code"
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

Next, [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) will be to plot the sparklines as shown below.

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## [8.2]{style="color:mediumvioletred"} Changing the pagesize

By default the pagesize is 10. In the code chunk below, arguments defaultPageSize is used to change the default setting.

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 180),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## [8.3]{style="color:mediumvioletred"} Adding points and labels

In the code chunk below `highlight_points` argument is used to show the **minimum** and **maximum** values points and `label` argument is used to label first and last values.

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## [8.4]{style="color:mediumvioletred"} Adding reference line

Use `statline` argument is used to show the **mean line.**

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 180),
    `Monthly Sales` = colDef(
      minWidth = 250,
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## [8.5]{style="color:mediumvioletred"} Adding bandline

Instead adding reference line, bandline can be added by using the `bandline` argument.

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## [8.6]{style="color:mediumvioletred"} Changing from sparkline to sparkbar

Instead of displaying the values as sparklines, we can display them as sparkbars as shiwn below.

```{r}
#| code-fold: true
#| code-summary: "Show code"
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

# [9]{style="color:mediumvioletred"} Reference

Kam, T. S. (2023, December 4). R for Visual Analytics. https://r4va.netlify.app/
